G\"{o}del's Ontological Proof of God's Existence (Draft)

### 2013-09-12 Natural Deduction/GodProof-ND.tex
Add draft definition of the natural deduction calculus

\usepackage{latexsym}
\usepackage{proof}
\usepackage{bussproofs}
\EnableBpAbbreviations
\newcommand{\rl}[1]{\RightLabel{#1}}

@@

\begin{calculus}
{The natural deduction calculus \ND}
{fig:ND}

@@

\s\s\s
\infer[\imp_I^n]{A \imp B}{ \infer*{B}{A^n} }
\s\s\s
\infer[\imp_E]{B}{A & A \imp B}

@@

\s
\infer[\all_I]{\all x. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x. A[x] }
\s\s
\infer[\ex_I]{\ex x. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\alpha]}{ \ex x. A[x] }

@@

ToDo: State eigen-variable conditions
\end{calculus}

\begin{calculus}
{\NDK: Rules for Modal Operators}
{fig:NDK}
\s\s\s
\infer[\nec_I]{\nec A}{A}
\s\s
\infer[\nec_E]{A}{\nec A}

@@

ToDo: state conditions
\end{calculus}

@@

ToDo: This section still needs more details. See Coq formalization for more details.

@@

\begin{theorem}
\label{T3}
Necessarily, God exists:

\noindent
\textbf{Theorem:} The existence of a God is necessary:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec \ex x. G(x)$} \RightLabel{$\nec$ I}
\UIC{$\nec\nec \ex x. G(x)$}\RightLabel{$\imp$ I, 1}
\UIC{$\nec \ex x. G(x)\imp\nec\nec \ex x. G(x)$} \RightLabel{propositional logic}
\UIC{$\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x)$}\RightLabel{$\nec$ I}
\UIC{$\nec(\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x))$}
\AXC{Axiom K} \dashedLine 
\BIC{$\nec\neg\nec\nec \ex x. G(x)\imp\nec\neg  \nec \ex x. G(x)$} \RightLabel{definition of $\pos$}
 \UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\pos\neg\nec \ex x. G(x)$}
\AXC{$ \Pi$} \dashedLine 
\UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}
 \BIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec$ E, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{$ $} \dashedLine  \RightLabel{Lemma \ref{L1}}
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp$ E}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec$ I}
\UIC{$\nec \neg\ex x. G(x)$}  \RightLabel{definition of $\pos$}
\UIC{$\neg \pos\ex x. G(x)$}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\AXC{ Corollary \ref{C1}} \dashedLine 
\UIC{$ \pos\ex x. G(x)$} \RightLabel{$\imp$ E}
 \BIC{$ \bot$} \RightLabel{$\imp$ I, 1}
 \UIC{$\neg\nec\pos\neg\nec \ex x. G(x)$}
\AXC{ Axiom B} \dashedLine 
\UIC{$ \neg\nec \ex x. G(x)\imp \nec\pos\neg\nec \ex x. G(x)$} \RightLabel{propositional logic}
\UIC{$ \neg \nec\pos\neg\nec \ex x. G(x)\imp  \nec \ex x. G(x)$} \RightLabel{$\imp$ E}
 \BIC{$ \nec \ex x. G(x)$}
\end{prooftree}
\end{small}

\textbf{Better proof, that directly proves $\ex x. G(x)$:}
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}

 \UIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec$ E, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{$ $} \dashedLine  \RightLabel{Lemma \ref{L1}}
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp$ E}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec$ I}
\UIC{$\nec \neg\ex x. G(x)$}  \RightLabel{definition of $\pos$}
\UIC{$\neg \pos\ex x. G(x)$} 
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\AXC{ Corollary \ref{C1}} \dashedLine 
\UIC{$ \pos\ex x. G(x)$} \RightLabel{$\imp$ E}
 \BIC{$ \bot$} \RightLabel{$\imp$ I, 1}
 \UIC{$\neg\nec\neg\nec \ex x. G(x)$} \RightLabel{definition of $\pos$}
 \UIC{$\neg\nec\pos \neg \ex x. G(x)$}
\AXC{ Axiom B} \dashedLine 
\UIC{$ \neg \ex x. G(x)\imp\nec\pos \neg \ex x. G(x)$} \RightLabel{propositional logic}
\UIC{$ \neg\nec\pos \neg \ex x. G(x)\imp  \neg  \neg \ex x. G(x)$} \RightLabel{$\imp$ E}
 \BIC{$ \neg  \neg \ex x. G(x)$}\RightLabel{$ \neg  \neg$ E}
 \UIC{$  \ex x. G(x)$}
\end{prooftree}
\end{small}

@@

The system used contains the $\nec$E-rule with the restriction that we have a $\nec$I below it. This is equivalent to modal sytem K that contains axiom K and necessitation rule N.  We aslo use axiom B ($ A\imp\nec\pos A$). No other modal axioms are needed.

@@

\begin{axiom}
\label{AK}
The transitivity axiom K. 
$$
\nec(A\imp B)\imp  (\nec A\imp \nec B)
$$
\end{axiom}

\begin{axiom}
\label{N}
The necessitation rule N.

If $A$ is a theorem, then $\nec A$ is a theorem. 
\end{axiom}

@@

\begin{small}
\begin{prooftree}
\AXC{$\nec(A\imp B)^2 $}\RightLabel{$\nec$ E}
\UIC{$A\imp B$} 
\AXC{$\nec A ^1 $}\RightLabel{$\nec$ E}
\UIC{$A$} \RightLabel{$\imp$ E}
 \BIC{$ B$} \RightLabel{$\nec$I}
\UIC{$ \nec B$} \RightLabel{$\imp$ I, 1}
 \UIC{$\nec A\imp \nec B$}  \RightLabel{$\imp$ I, 2}
 \UIC{$\nec(A\imp B)\imp  (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$A^1 $}\dashedLine
\UIC{$ B$} \RightLabel{$\imp$ I, 1}
\UIC{$A\imp B$} \RightLabel{rule N}
\UIC{$\nec(A\imp B)$} 
\AXC{Axiom K}\dashedLine
\UIC{$\nec(A\imp B)\imp  (\nec A\imp \nec B)$} \RightLabel{$\imp$ E}
\BIC{$\nec A\imp \nec B$} 
\AXC{$\nec A ^1 $}\RightLabel{$\imp$ E}
\BIC{$\nec B$}
\end{prooftree}
\end{small}

>>>>

\usepackage{fancybox}
\usepackage{latexsym}
\usepackage{proof}
\usepackage{bussproofs}
\EnableBpAbbreviations
\newcommand{\rl}[1]{\RightLabel{#1}}

@@

\begin{calculus}
{The intuitionistic natural deduction calculus \ND}
{fig:ND}

@@

\s\s
\infer[\imp_I]{A \imp B}{ B }
\s\s
\infer[\imp_I^n]{A \imp B}{ \infer*{B}{\infer[n]{A}{}} }
\s\s
\infer[\imp_E]{B}{A & A \imp B}

@@

\s\s
\infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
\s\s
\infer[\vee_{I_1}]{A \vee B}{A}
\s\s
\infer[\vee_{I_2}]{A \vee B}{B}

\s
\infer[\all_I]{\all x. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x. A[x] }
\s\s
\infer[\ex_I]{\ex x. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\beta]}{ \ex x. A[x] }

@@

ToDo: add intuitionistic rule for negation/bottom
\end{calculus}

\begin{calculus}
{Classical Rules/Axioms}
{fig:Classical}

ToDo: add classical rules/axioms that we need
\end{calculus}

\begin{calculus}
{Rules for Modal Operators}
{fig:NDK}

@@

\s\s\s\s
\infer[\nec_I]{\nec A}{\fbox{\infer*{A}{} }}
\s\s\s
\infer[\nec_E]{\fbox{ \infer*{}{A} }  }{\nec A}

\vspace{1em}
\end{calculus}

\noindent
A \emph{derivation} is a directed acyclich graph whose nodes are formulas and whose edges correspond to applications of the inference rules shown in Figures \ref{fig:ND} and \ref{fig:NDK}. Parts of a derivation may be surrounded by boxes. A \emph{proof} is a derivation that additionally satisfies the following conditions:

\begin{itemize}
\item \textbf{eigen-variable conditions:}
if $\rho$ is a $\all_I$ inference eliminating a variable $\alpha$, then any occurrence of $\alpha$ in the proof should be an ancestor of the occurrence of $\alpha$ eliminated by $\rho$;
if $\rho$ is a $\ex_E$ inference introducing a variable $\beta$, then any occurrence of $\beta$ in the proof should be a descendant of the occurrence of $\beta$ introduced by $\rho$.
%
\item \textbf{boxed assumption condition:} any assumption should be discharged within the box where it is made.
%
\item \textbf{unboxed root condition:} the proof's root should not be inside any box.
\end{itemize}

\noindent
Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to a proof shown elsewhere. Dotted lines are used to indicate folding and unfolding of definitions.

@@

@@

\begin{theorem}
\label{T3}
Necessarily, God exists:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec \ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec\nec \ex x. G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\nec \ex x. G(x)\imp\nec\nec \ex x. G(x)$} \RightLabel{propositional logic}
\UIC{$\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x)$}\RightLabel{$\nec_I$}
\UIC{$\nec(\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x))$}
\AXC{Axiom K} \dashedLine 
\BIC{$\nec\neg\nec\nec \ex x. G(x)\imp\nec\neg  \nec \ex x. G(x)$} \dottedLine \RightLabel{$\pos$}
 \UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\pos\neg\nec \ex x. G(x)$}
\AXC{$ \Pi$} \dashedLine 
\UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}
 \BIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec_E$, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{$ $} \dashedLine  \RightLabel{Lemma \ref{L1}}
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp_E$}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \neg\ex x. G(x)$}  \dottedLine
\UIC{$\neg \pos\ex x. G(x)$}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\AXC{ Corollary \ref{C1}} \dashedLine 
\UIC{$ \pos\ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \bot$} \RightLabel{$\imp_I^1$}
 \UIC{$\neg\nec\pos\neg\nec \ex x. G(x)$}
\AXC{ Axiom B} \dashedLine 
\UIC{$ \neg\nec \ex x. G(x)\imp \nec\pos\neg\nec \ex x. G(x)$} \doubleLine
\UIC{$ \neg \nec\pos\neg\nec \ex x. G(x)\imp  \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \nec \ex x. G(x)$}
\end{prooftree}
\end{small}

\textbf{Better proof, that directly proves $\ex x. G(x)$:}
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec_E$, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{ Lemma \ref{L1} } \dashedLine
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp_E$}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \neg\ex x. G(x)$}  \dottedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\AXC{ Corollary \ref{C1}} \dashedLine 
\UIC{$ \pos\ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \bot$} \RightLabel{$\imp_I^1$}
 \UIC{$\neg\nec\neg\nec \ex x. G(x)$} \dottedLine
 \UIC{$\neg\nec\pos \neg \ex x. G(x)$}
\AXC{ Axiom B} \dashedLine 
\UIC{$ \neg \ex x. G(x)\imp\nec\pos \neg \ex x. G(x)$} \doubleLine
\UIC{$ \neg\nec\pos \neg \ex x. G(x)\imp  \neg  \neg \ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \neg  \neg \ex x. G(x)$}\RightLabel{$ \neg  \neg$ E}
 \UIC{$  \ex x. G(x)$}
\end{prooftree}
\end{small}

@@

The system used contains the $\nec_E$-rule with the restriction that we have a $\nec_I$ below it. This is equivalent to modal sytem K that contains axiom K and necessitation rule N.  We aslo use axiom B ($ A\imp\nec\pos A$). No other modal axioms are needed.

@@

\begin{axiom}[The transitivity axiom K]
\label{K}
$$
\nec(A\imp B)\imp  (\nec A\imp \nec B)
$$
\end{axiom}

\begin{axiom}[The necessitation rule]
\label{Necessitation}
If $A$ is a theorem, then $\nec A$ is a theorem. 
\end{axiom}

@@

\begin{small}
\begin{prooftree}
\AXC{$\nec(A\imp B)^2 $}\RightLabel{$\nec_E$}
\UIC{$A\imp B$} 
      \AXC{$\nec A ^1 $}\RightLabel{$\nec_E$}
      \UIC{$A$} \RightLabel{$\imp_E$}
   \BIC{$ B$} \RightLabel{$\nec_I$}
   \UIC{$ \nec B$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec B$}  \RightLabel{$\imp_I^2$}
   \UIC{$\nec(A\imp B)\imp  (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$A^1 $}\noLine
\UIC{$\vdots$}\noLine
\UIC{$ B$} \RightLabel{$\imp_I^1$}
\UIC{$A\imp B$} \RightLabel{Necessitation}
\UIC{$\nec(A\imp B)$} 
      \AXC{Axiom K}\dashedLine
      \UIC{$\nec(A\imp B)\imp  (\nec A\imp \nec B)$} \RightLabel{$\imp_E$}
  \BIC{$\nec A\imp \nec B$} 
        \AXC{$\nec A ^1 $}\RightLabel{$\imp_E$}
    \BIC{$\nec B$}
\end{prooftree}
\end{small}
